What are the contributing factors to poor mental health in U.S?

### 2022-03-21 outputs/paper/paper.Rmd
Updated Paper.rmd with cross referencing and supplemental survey

---
title: "Age and gender in the 2021 US GSS"
subtitle: "Possible selection effect in ages 18-35"
author: 
  - Ray Wen, Isfandyar Virani, Rayhan Walia
thanks: "Code and data are available at: https://github.com/ray0130/usa_gss."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

####Workplace Setup####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(haven)

raw_data <- haven::read_dta("../../inputs/data/2021_stata/gss2021.dta")

gss2021 <- raw_data
```

@@

Mental health is a growing problem in the world, with a 13% worldwide increase in the last decade ([cite]), which makes this an important point for policymakers in government. During the last two years, the COVID-19 pandemic has not helped the situation either, causing a dramatic increase in anxiety and depression across the United States ([cite]).

In this report, we monitor the mental health status of adults (over 18) in the United States, through the COVID-19 pandemic. We observe various groups (of sex, sexual orientation, social and income class) and how they are affected by mental health. This report provides an understanding for extreme cases such as the pandemic, and via collecting as much data as possible, a solution could be made to combat this problem. The effect of the pandemic on the survey itself is also noticeable, with the dramatic decrease in response rate (over 50% before the pandemic to 17.4% for GSS 2021) due to the shift to the web-based system. This can be easily rectified in the future (with the lack of COVID-19), but not much can be done in an extreme case such as the pandemic.

@@

To get better insights on what factors affect Mental health for different parts of the population, we utilized the 2021 US General Social Survey [cite] from NORC at the University of Chicago [CITE]. The raw dataset is a Stata file and was imported in R using the package Haven. It includes data from 568 Survey Questions for 4032 Survey Participants. Using the R package [CITE] tidyverse and dplyr [cite], we were able to clean and perform exploratory data analysis on the dataset to get insights into the data. Further, we used R package ggplot2 [CITE] to visualize the data for this paper.

The survey of interest in this report is the General Social Survey (GSS [CITE GSS]), one that monitors pubic opinion and behavior in the United States. It has been conducted since 1972 by the NORC at the University of Chicago and funded by the National Science Foundation(NSF); aiming to minimize all changes via retaining similar sampling and questioning approaches. [MAYBE A LITTLE MORE ON SURVEY]
In each round, the GSS contains a set of repeating modules and a section of topical modules that is subject to change in every round and each participant will be given a subset of repeating modules and topical modules (it may not contain all modules).

## Sample
[TABLE 1 GSS CODEBOOK. TALK ABOUT ASPECTS]

@@

The GSS has further made changes to the survey response options as a result of the shift from in-person interviews to web and phone-based interviews. Originally the survey contains “volunteered responses” that are not displayed to the responders and is used by the interviewer. An example would be the option “about right” in response to the question “In general, do you think the courts in this area deal too harshly or not harshly enough with criminals?” (Cite GSS?) However, in this year's survey, the interviewers are unable to add these volunteered responses due to the lack of interaction on the web and phone-based interviews. Therefore, the GSS adopted a new structure for these volunteered responses on the web mode by creating two different versions of forms with one containing the volunteered responses as an option and the other without. Thus, any changes in the public opinion seen in the 2021 GSS data could be due to either actual changes in public opinion or the result of this newly adopted methodology and should be closely monitored and considered in analysis on the survey.

@@

Using a number of responses through the survey, we are able to observe the results visually using ggplot [CITE GG]; aiding in the understanding of the results.

@@

```{r fig1, warning=FALSE, fig.cap="Mental Health - Male Vs Female", message = FALSE, echo=FALSE}
labels <- c("Female", "Male")
gss2021$sex_group <- cut(gss2021$sex, 2, labels = labels)

#Recoding mental health
# Changed good or better to 'Good'
# Change poor and fair to 'Not Good'
gss2021 <- gss2021 %>%
  mutate(`Mental Health` = case_when(
    hlthmntl == 1 ~ 'Good',
    hlthmntl == 2 ~ 'Good',
    hlthmntl == 3 ~ 'Good',
    hlthmntl == 4 ~ 'Not Good',
    hlthmntl == 5 ~ 'Not Good',
  ))

#=============================================================================
# Percent Graph - Mental health - Male Vs Female
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sex_group)) %>% ggplot(aes(x= `Mental Health`,  group=sex_group)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x = 'Mental Health', title = 'Mental Health: Male vs. Female') +
  facet_grid(~sex_group) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

(Figure 1) – Mental health (Male vs Female)) shows that 18.3% of the male respondents reported having ‘Not Good’ mental health and 81.7% reported ‘Good’ mental health compared to 13.8% of the female respondents reported having ‘Not Good’ mental health and 86.2% reported ‘Good’ mental health.

@@

```{r fig2, warning=FALSE, fig.cap="Mental Health – Physical Health", message = FALSE, echo=FALSE}

#Recoding physical health
gss2021 <- gss2021 %>%
  mutate(`Physical Health` = case_when(
    hlthphys == 1 ~ 'Good',
    hlthphys == 2 ~ 'Good',
    hlthphys == 3 ~ 'Good',
    hlthphys == 4 ~ 'Not Good',
    hlthphys == 5 ~ 'Not Good',
  ))

#=============================================================================

# Percent Graph Physical health - Good Mental health vs Bad Mental Health
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sex_group), !is.na(`Physical Health`)) %>%
  ggplot(aes(x= `Physical Health`,  group= `Mental Health`)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", fill="Physical Health", x = 'Physical Health', title = "Mental Health vs. Physical Health" ) +
  facet_grid(~`Mental Health` , labeller = label_both) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

In (Figure 2), we can see that for people who reported ‘Not Good’ mental health, 56% reported having ‘Not Good’ physical health and 44% reported ‘Good’ physical health. While, for the people who reported ‘Good’ mental health, 13% reported ‘Not Good’ physical health, and 87% reported ‘Good’ physical health.

@@

```{r fig3, warning=FALSE, fig.cap="Physical Health Comparison of Male Vs. Female with Poor Mental Health", message = FALSE, echo=FALSE}

# Percent Graph - Poor Mental health = Physical health within Sex_Group comparison
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sex_group), !is.na(`Physical Health`), `Mental Health` =='Not Good') %>%
  ggplot(aes(x= `Physical Health`,  group=sex_group)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x="Physical Health", title = "Physical Health Comparison of Male Vs. Female with Poor Mental Health") +
  facet_grid(~sex_group) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

In Figure 3, we found that for participants whose mental health is ‘Not Good’, both Male and Female participants have similar distribution of physical health at 54.5% Female and 56.4% Male reporting ‘Not Good’ physical health and 45.5% of Female  43.6% of Male reporting ‘Good’ physical health while having poor mental health.

This displays a pattern, that males and females have nearly identical percentages for their physical health; given poor mental health. [APPENDIX - COMPARE MALE FEMALE PHYS HEALTH WITH “GOOD” MENTAL HEALTH]

@@

```{r fig4, warning=FALSE, fig.cap="Mental Health – Age Groups", message = FALSE, echo=FALSE}

# Recording Age Group
gss2021 <- gss2021 %>%
  mutate(`Age Group` = case_when(
    age >= 18 & age <= 29 ~ '18-29',
    age >= 30 & age <=39~ '30-39',
    age >= 40 & age <=49~ '40-49',
    age >= 50 & age <=64~ '50-64',
    age >= 65 & age <=89~ '65-89'
  ))

#=============================================================================

# Percent Graph - Mental health (Age Group Analysis)
gss2021 %>% filter(!is.na(`Mental Health`),!is.na(`Age Group`)) %>%
  ggplot(aes(x= `Mental Health`,  group=`Age Group`)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x="Mental Health", title = "Mental Health: Comparison of Different Age Groups") +
  facet_grid(~`Age Group`, labeller = label_both) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

In Figure 4, we can see that the Age Group of 18-29 has the highest ‘Not Good’ mental health at 34.5% followed by Ages 30-39 with 23.5%, 16.4% for Age Group 40-49, 13.4% for Age Group 50-64, and 9.1% for Age group 65-89.

Here, we can observe a very interesting pattern, that quality of mental health is highly dependent on one’s age; with the youngest respondents reporting the poorest mental health.  [CITE]

@@

```{r fig5, warning=FALSE, fig.cap="Mental Health – Sexual Orientation", message = FALSE, echo=FALSE}

# Recoding Sexual orientation

gss2021 <- gss2021 %>%
  mutate(sexual_orientation = case_when(
    sexornt == 1 ~ 'LGBTQ+',
    sexornt == 2 ~ 'LGBTQ+',
    sexornt == 3 ~ 'Heterosexual'
  ))

#=============================================================================

# Percent Graph - Mental health (Sexual Orientation Analysis)
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sexual_orientation)) %>%
  ggplot(aes(x= `Mental Health`,  group=sexual_orientation)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x="Mental Health", title = "Mental Health: Heterosexual vs. LGBTQ+") +
  facet_grid(~sexual_orientation) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

In Figure 5, we can see that 35% of LGBTQ+ reported their mental health as being Not Good (I.e, Poor or Fair) compared to 14% of Heterosexual. LGBTQ+ population experience a higher rate of poor mental health compared to the Heterosexual population.

@@

```{r fig6, warning=FALSE, fig.cap="Mental Health – Social Class", message = FALSE, echo=FALSE}

gss2021 <- gss2021 %>%
  mutate(`Social Class` = case_when(
    class1 == 1 ~ "lower",
    class1 == 2 ~ "working",
    class1 == 3 ~ "lower middle",
    class1 == 4 ~ "middle",
    class1 == 5 ~ "upper middle",
    class1 == 6 ~ "upper"))

#=============================================================================
# 'Mental Health: Comparison of Different Social Classes'
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(`Social Class`)) %>% ggplot(aes(x= `Mental Health`,  group= `Social Class`))+
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", fill="mntl_poor", title = 'Mental Health: Comparison of Different Social Classes') +
  facet_grid(~factor(`Social Class`, levels = c('lower', 'working','lower middle', 'middle', 'upper middle', 'upper'))) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

Figure 6 shows a bar graph of the mental health rating for each of the social classes. These social classes are self-identified by the responder to see where they see fit in the 6 different social classes.  44.2% of the lower social class, 21.3% of the working class, 20% of the lower middle class, 12.6% of the middle class, 7.5% of the upper-middle class, and 27.3% of the upper-class respondents responded with poor mental health.

@@

```{r fig7, warning=FALSE, fig.cap="Mental Health – Total Family Income", message = FALSE, echo=FALSE}
# Recoding Family Income
gss2021 <- gss2021 %>% 
  mutate(Income = case_when(
    income16 <= 8 ~ "< $10,000",
    income16 <= 12 ~ "$10,000 - $20,000",
    income16 <= 17 ~ "$20,000 - $40,000",
    income16 <= 19 ~ "$40,000 - $60,000",
    income16 <= 21 ~ "$60,000 - $90,000",
    income16 > 21 ~ "> $90,000"))

#=============================================================================
# 'Mental Health: Comparison by Total Family Income'
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(Income)) %>%  ggplot(aes(x= `Mental Health`,  group=Income))+
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent" , title = 'Mental Health: Comparison by Total Family Income') +
  facet_grid(~factor(Income, levels = c('< $10,000', "$10,000 - $20,000", "$20,000 - $40,000", "$40,000 - $60,000", "$60,000 - $90,000", "> $90,000"))) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

Figure 7  shows a bar graph of the mental health comparison for different family income groups. 32.3% in the lowest income bracket (< \$10,000), 29.5%  in the lower-income (\$10,000 - \$20,000), 23.8% in lower middle income (\$20,000 - \$40,000), 18.3% in the middle income (\$40,000 - \$60,000), 13.4% in the upper middle income (\$60,000 - \$90,000), and 7.7% in the upper income (> \$90,000)  responded with ‘Not Good’ mental health. There is a clear trend in lower income family income having worse mental health compared to higher income groups.

@@

```{r fig8, warning=FALSE, fig.cap="Mental Health – Age When First Child was Born", message = FALSE, echo=FALSE}
# Recoding Age when first kid is born
gss2021 <- gss2021 %>%
  mutate(`Kid Born` = case_when(
    agekdbrn < 18 ~ '< 18',
    agekdbrn >= 18 & agekdbrn <= 29 ~ '18-29',
    agekdbrn >= 30 & agekdbrn <=39~ '30-39',
    agekdbrn >= 40 & agekdbrn <=49~ '40-49',
    agekdbrn >= 50 & agekdbrn <=64~ '50-64',
    agekdbrn >= 65 & agekdbrn <=89~ '65-89'
  ))

#=============================================================================
#  "Mental Health: Comparison by age when first child was born"
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(`Kid Born`))  %>% ggplot(aes(x= `Mental Health`,  group=`Kid Born`))+
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", title = "Mental Health: Comparison by Age When First Child was Born") +
  facet_grid(~`Kid Born`) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")

```

Figure 8  shows a bar graph of the respondent’s age when they had their firstborn child. 18.9% of age younger than 18, 14.4% of age between 18 and 29, 10.2% of age between 30-39, 7% of age between 40-49, and 0% of age 50 and above responded with poor mental health.

@@

Due to the COVID-19 pandemic, there has been a rise in mental health ([cite]). With the rise in COVID-19 cases and social distancing being enforced by the governments, there is an increase in anxiety and depression due to isolation, job loss, and fear of infection [cite]. Looking into our data analysis we find some key insights that can help policymakers to take appropriate actions to reduce mental health suffering which can reduce long-term social and economical costs for society.

@@

Although the difference between ‘Not Good’ mental health between female and male participants isn’t huge (less than 5%) in Figure 1, our findings are a contradiction to studies such as (Cite) and (Cite) that suggests that females are more likely than males to report worse mental health, especially since physical distancing began due to the pandemic.

@@

In (Figure 2), we see those survey participants who reported their mental health as Poor or Fair have significantly worse physical health compared to participants who have good mental health. This finding can also be seen in different studies such as (cite) and is mentioned (cite) & (cite) as physical health and mental health are inextricably related. Physical health can be managed by exercising, eating well, and avoiding smoking & drinking.

@@

LGBTQ+ individuals have poor mental health compared to heterosexual individuals (cite). This is due to many factors such as lack of social support which has an association with higher levels of depression, anxiety, and low self-esteem (cite)

>>>>

---
title: "What are the contributing factors to poor mental health in U.S?"
subtitle: "Analysis of US General Social Survey of 2021"
author: 
  - Ray Wen, Isfandyar Virani, Rayhan Walia
thanks: "Code and data are available at: https://github.com/ray0130/usa_gss."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "In this paper, the 2021 US General Social Survey was analyzed to examine how mental health affects different parts of the population during the COVID-19 pandemic. The analysis was able to determine that Male participants, participants with poor physical health, Younger Age Groups between 18 and 29, LGBTQ+, participants with lower income and low social class, and participants that had their firstborn child at a younger age have poor mental health compared to their counterparts. This is an important result as it can help policymakers to take appropriate actions to reduce mental health suffering which can in-turn reduce long-term social and economical costs for society."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
####Workplace Setup####
library(tidyverse)
library(ggplot2)
library(dplyr)
library(haven)
raw_data <- haven::read_dta("../../inputs/data/2021_stata/gss2021.dta")
gss2021 <- raw_data
```

@@

Mental health is a growing problem in the world, with a 13% worldwide increase in the last decade [@worldhealthorganization], which makes this an important point for policymakers in government. During the last two years, the COVID-19 pandemic has not helped the situation either, causing a dramatic increase in anxiety and depression across the United States [@mentalhealthamerica].

In this report, we monitor the mental health status of adults (over 18) in the United States, through the COVID-19 pandemic. We observe various groups (of sex, sexual orientation, social and income class) and how they are affected by mental health. This report provides an understanding of extreme cases such as the pandemic, and via collecting as much data as possible, a solution could be made to combat this problem. The effect of the pandemic on the survey itself is also noticeable, with the dramatic decrease in response rate (over 50% before the pandemic to 17.4% for GSS 2021) due to the shift to the web-based system. This can be easily rectified in the future (with the lack of COVID-19), but not much can be done in an extreme case such as the pandemic.

@@

To get better insights on what factors affect Mental health for different parts of the population, we utilized the 2021 US General Social Survey [@norc] from NORC at the University of Chicago [@norc-at-the-university-of-chicago]. The raw dataset is a Stata file and was imported in R using the package Haven. It includes data from 568 Survey Questions for 4032 Survey Participants. Using the R [@citeR] package tidyverse [@citetidyverse] and dplyr [@dplyr], we were able to clean and perform exploratory data analysis on the dataset to get insights into the data. Further, we used R package ggplot2 [@citeggplot2] to visualize the data for this paper.

The survey of interest in this report is the General Social Survey [@norc], one that monitors pubic opinion and behavior in the United States. It has been conducted since 1972 by the NORC at the University of Chicago and funded by the National Science Foundation(NSF); aiming to minimize all changes via retaining similar sampling and questioning approaches.
In each round, the GSS contains a set of repeating modules and a section of topical modules that is subject to change in every round and each participant will be given a subset of repeating modules and topical modules (it may not contain all modules).

## Sample

@@

The GSS has further made changes to the survey response options as a result of the shift from in-person interviews to web and phone-based interviews. Originally the survey contains “volunteered responses” that are not displayed to the responders and is used by the interviewer. An example would be the option “about right” in response to the question “In general, do you think the courts in this area deal too harshly or not harshly enough with criminals?” ([@norc]). However, in this year's survey, the interviewers are unable to add these volunteered responses due to the lack of interaction on the web and phone-based interviews. Therefore, the GSS adopted a new structure for these volunteered responses on the web mode by creating two different versions of forms with one containing the volunteered responses as an option and the other without. Thus, any changes in the public opinion seen in the 2021 GSS data could be due to either actual changes in public opinion or the result of this newly adopted methodology and should be closely monitored and considered in analysis on the survey.

@@

Using a number of responses through the survey, we are able to observe the results visually using ggplot [@citeggplot2]; aiding in the understanding of the results.

@@

```{r fig1, warning=FALSE, fig.cap="Mental Health - Male Vs Female", message = FALSE, echo=FALSE, fig.width=5, fig.height=5}
labels <- c("Female", "Male")
gss2021$sex_group <- cut(gss2021$sex, 2, labels = labels)
#Recoding mental health
# Changed good or better to 'Good'
# Change poor and fair to 'Not Good'
gss2021 <- gss2021 %>%
  mutate(`Mental Health` = case_when(
    hlthmntl == 1 ~ 'Good',
    hlthmntl == 2 ~ 'Good',
    hlthmntl == 3 ~ 'Good',
    hlthmntl == 4 ~ 'Not Good',
    hlthmntl == 5 ~ 'Not Good',
  ))
#=============================================================================
# Percent Graph - Mental health - Male Vs Female
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sex_group)) %>% ggplot(aes(x= `Mental Health`,  group=sex_group)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x = 'Mental Health', title = 'Mental Health: Male vs. Female') +
  facet_grid(~sex_group) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

(Figure \@ref(fig:fig1)) – Mental health (Male vs Female)) shows that 18.3% of the male respondents reported having ‘Not Good’ mental health and 81.7% reported ‘Good’ mental health compared to 13.8% of the female respondents reported having ‘Not Good’ mental health and 86.2% reported ‘Good’ mental health.

@@

```{r fig2, warning=FALSE, fig.cap="Mental Health – Physical Health", message = FALSE, echo=FALSE, fig.width=5, fig.height=5}
#Recoding physical health
gss2021 <- gss2021 %>%
  mutate(`Physical Health` = case_when(
    hlthphys == 1 ~ 'Good',
    hlthphys == 2 ~ 'Good',
    hlthphys == 3 ~ 'Good',
    hlthphys == 4 ~ 'Not Good',
    hlthphys == 5 ~ 'Not Good',
  ))
#=============================================================================
# Percent Graph Physical health - Good Mental health vs Bad Mental Health
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sex_group), !is.na(`Physical Health`)) %>%
  ggplot(aes(x= `Physical Health`,  group= `Mental Health`)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", fill="Physical Health", x = 'Physical Health', title = "Mental Health vs. Physical Health" ) +
  facet_grid(~`Mental Health` , labeller = label_both) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

In (Figure \@ref(fig:fig2)), we can see that for people who reported ‘Not Good’ mental health, 56% reported having ‘Not Good’ physical health and 44% reported ‘Good’ physical health. While, for the people who reported ‘Good’ mental health, 13% reported ‘Not Good’ physical health, and 87% reported ‘Good’ physical health.

@@

```{r fig3, warning=FALSE, fig.cap="Physical Health Comparison of Male Vs. Female with Poor Mental Health", message = FALSE, echo=FALSE, fig.width=7, fig.height=6}
# Percent Graph - Poor Mental health = Physical health within Sex_Group comparison
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sex_group), !is.na(`Physical Health`), `Mental Health` =='Not Good') %>%
  ggplot(aes(x= `Physical Health`,  group=sex_group)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x="Physical Health", title = "Physical Health Comparison of Male Vs. Female with Poor Mental Health") +
  facet_grid(~sex_group) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

In (Figure \@ref(fig:fig3)), we found that for participants whose mental health is ‘Not Good’, both Male and Female participants have similar distribution of physical health at 54.5% Female and 56.4% Male reporting ‘Not Good’ physical health and 45.5% of Female  43.6% of Male reporting ‘Good’ physical health while having poor mental health.

This displays a pattern, that males and females have nearly identical percentages for their physical health; given poor mental health.

@@

```{r fig4, warning=FALSE, fig.cap="Mental Health – Age Groups", message = FALSE, echo=FALSE, fig.width=7, fig.height=5}
# Recording Age Group
gss2021 <- gss2021 %>%
  mutate(`Age Group` = case_when(
    age >= 18 & age <= 29 ~ '18-29',
    age >= 30 & age <=39~ '30-39',
    age >= 40 & age <=49~ '40-49',
    age >= 50 & age <=64~ '50-64',
    age >= 65 & age <=89~ '65-89'
  ))
#=============================================================================
# Percent Graph - Mental health (Age Group Analysis)
gss2021 %>% filter(!is.na(`Mental Health`),!is.na(`Age Group`)) %>%
  ggplot(aes(x= `Mental Health`,  group=`Age Group`)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x="Mental Health", title = "Mental Health: Comparison of Different Age Groups") +
  facet_grid(~`Age Group`, labeller = label_both) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

In (Figure \@ref(fig:fig4)), we can see that the Age Group of 18-29 has the highest ‘Not Good’ mental health at 34.5% followed by Ages 30-39 with 23.5%, 16.4% for Age Group 40-49, 13.4% for Age Group 50-64, and 9.1% for Age group 65-89.

Here, we can observe a very interesting pattern, that quality of mental health is highly dependent on one’s age; with the youngest respondents reporting the poorest mental health.

@@

```{r fig5, warning=FALSE, fig.cap="Mental Health – Sexual Orientation", message = FALSE, echo=FALSE, fig.width=5, fig.height=5}
# Recoding Sexual orientation
gss2021 <- gss2021 %>%
  mutate(sexual_orientation = case_when(
    sexornt == 1 ~ 'LGBTQ+',
    sexornt == 2 ~ 'LGBTQ+',
    sexornt == 3 ~ 'Heterosexual'
  ))
#=============================================================================
# Percent Graph - Mental health (Sexual Orientation Analysis)
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(sexual_orientation)) %>%
  ggplot(aes(x= `Mental Health`,  group=sexual_orientation)) +
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", x="Mental Health", title = "Mental Health: Heterosexual vs. LGBTQ+") +
  facet_grid(~sexual_orientation) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

In (Figure \@ref(fig:fig5)), we can see that 35% of LGBTQ+ reported their mental health as being Not Good (I.e, Poor or Fair) compared to 14% of Heterosexual. LGBTQ+ population experience a higher rate of poor mental health compared to the Heterosexual population.

@@

```{r fig6, warning=FALSE, fig.cap="Mental Health – Social Class", message = FALSE, echo=FALSE, fig.width=9, fig.height=5}
gss2021 <- gss2021 %>%
  mutate(`Social Class` = case_when(
    class1 == 1 ~ "lower",
    class1 == 2 ~ "working",
    class1 == 3 ~ "lower middle",
    class1 == 4 ~ "middle",
    class1 == 5 ~ "upper middle",
    class1 == 6 ~ "upper"))
#=============================================================================
# 'Mental Health: Comparison of Different Social Classes'
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(`Social Class`)) %>% ggplot(aes(x= `Mental Health`,  group= `Social Class`))+
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", fill="mntl_poor", title = 'Mental Health: Comparison of Different Social Classes') +
  facet_grid(~factor(`Social Class`, levels = c('lower', 'working','lower middle', 'middle', 'upper middle', 'upper'))) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

(Figure \@ref(fig:fig6)) shows a bar graph of the mental health rating for each of the social classes. These social classes are self-identified by the responder to see where they see fit in the 6 different social classes.  44.2% of the lower social class, 21.3% of the working class, 20% of the lower middle class, 12.6% of the middle class, 7.5% of the upper-middle class, and 27.3% of the upper-class respondents responded with poor mental health.

@@

```{r fig7, warning=FALSE, fig.cap="Mental Health – Total Family Income", message = FALSE, echo=FALSE, fig.width=8, fig.height=5}
# Recoding Family Income
gss2021 <- gss2021 %>% 
  mutate(Income = case_when(
    income16 <= 8 ~ "< $10,000",
    income16 <= 12 ~ "$10,000 - $20,000",
    income16 <= 17 ~ "$20,000 - $40,000",
    income16 <= 19 ~ "$40,000 - $60,000",
    income16 <= 21 ~ "$60,000 - $90,000",
    income16 > 21 ~ "> $90,000"))
#=============================================================================
# 'Mental Health: Comparison by Total Family Income'
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(Income)) %>%  ggplot(aes(x= `Mental Health`,  group=Income))+
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent" , title = 'Mental Health: Comparison by Total Family Income') +
  facet_grid(~factor(Income, levels = c('< $10,000', "$10,000 - $20,000", "$20,000 - $40,000", "$40,000 - $60,000", "$60,000 - $90,000", "> $90,000"))) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

(Figure \@ref(fig:fig7)) shows a bar graph of the mental health comparison for different family income groups. 32.3% in the lowest income bracket (< \$10,000), 29.5%  in the lower-income (\$10,000 - \$20,000), 23.8% in lower middle income (\$20,000 - \$40,000), 18.3% in the middle income (\$40,000 - \$60,000), 13.4% in the upper middle income (\$60,000 - \$90,000), and 7.7% in the upper income (> \$90,000)  responded with ‘Not Good’ mental health. There is a clear trend in lower income family income having worse mental health compared to higher income groups.

@@

```{r fig8, warning=FALSE, fig.cap="Mental Health – Age When First Child was Born", message = FALSE, echo=FALSE, fig.width=8, fig.height=5}
# Recoding Age when first kid is born
gss2021 <- gss2021 %>%
  mutate(`Kid Born` = case_when(
    agekdbrn < 18 ~ '< 18',
    agekdbrn >= 18 & agekdbrn <= 29 ~ '18-29',
    agekdbrn >= 30 & agekdbrn <=39~ '30-39',
    agekdbrn >= 40 & agekdbrn <=49~ '40-49',
    agekdbrn >= 50 & agekdbrn <=64~ '50-64',
    agekdbrn >= 65 & agekdbrn <=89~ '65-89'
  ))
#=============================================================================
#  "Mental Health: Comparison by age when first child was born"
gss2021 %>% filter(!is.na(`Mental Health`), !is.na(`Kid Born`))  %>% ggplot(aes(x= `Mental Health`,  group=`Kid Born`))+
  geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
  geom_text(aes( label = scales::percent(..prop..),
                 y= ..prop.. ), stat= "count", vjust = -.5) +
  labs(y = "Percent", title = "Mental Health: Comparison by Age When First Child was Born") +
  facet_grid(~`Kid Born`) +
  scale_y_continuous(labels = scales::percent) + theme(legend.position="none") + scale_fill_brewer(palette = "Spectral")
```

(Figure \@ref(fig:fig8)) shows a bar graph of the respondent’s age when they had their firstborn child. 18.9% of age younger than 18, 14.4% of age between 18 and 29, 10.2% of age between 30-39, 7% of age between 40-49, and 0% of age 50 and above responded with poor mental health.

@@

Due to the COVID-19 pandemic, there has been a rise in mental health [@katekelland_2020]. With the rise in COVID-19 cases and social distancing being enforced by the governments, there is an increase in anxiety and depression due to isolation, job loss, and fear of infection [@world-health-organization]. Looking into our data analysis we find some key insights that can help policymakers to take appropriate actions to reduce mental health suffering which can reduce long-term social and economical costs for society.

@@

Although the difference between ‘Not Good’ mental health between female and male participants isn’t huge (less than 5%) in (Figure \@ref(fig:fig1)), our findings are a contradiction to studies such as [@Otten2021] and [@moyser_2020] that suggests that females are more likely than males to report worse mental health, especially since physical distancing began due to the pandemic.

@@

In (Figure \@ref(fig:fig2)), we see those survey participants who reported their mental health as Poor or Fair have significantly worse physical health compared to participants who have good mental health. This finding can also be seen in different studies such as [@Vaillant1979] and is mentioned [@mental-health-foundation_2022] & [@cmha-ontario] as physical health and mental health are inextricably related. Physical health can be managed by exercising, eating well, and avoiding smoking & drinking.

@@

LGBTQ+ individuals have poor mental health compared to heterosexual individuals [@McDonald2018]. This is due to many factors such as lack of social support which has an association with higher levels of depression, anxiety, and low self-esteem [@McDonald2018].
